### 2017-11-20 paper/response_to_reviewers.tex

\item{\textcolor{blue}{The package also employs a rarefaction function, but rarefaction can be problematic in the sense that rarefaction curves can sometimes cross such that the sampling level used to make comparisons can effect which of two or more subsets is considered the most disparate. This is why algorithms designed to sample "fairly" rather than equally have been explored, see for example Kotric and Knoll (2015, Paleobiology, 41, 68–). A user may also wish to rarefy by some external data source, such as number of occurrences, or geographic regions. These are obviously major additions, but are something the author might wish to add in future.}}

This is again an excellent suggestion!
I will develop more complex rarefaction algorithms based on sampling and occurrence based on methods applied empirically by Kotric and Knoll 2015 \textit{Palaeobiology} and Butler et al 2012 \textit{Evolution}.
I've added a note

% %TG: TODO maybe!

\item{\textcolor{blue}{Other minor additions the author may also wish to consider are the permutation test from Brusatte et al. (2014, Current Biology, 24, 2386-), a means of adding the geologic timescale to time series plots (e.g., using the geoscale package)}}

@@

See Reviewer 1 minor comment \ref{abstract_typo}.

\item{\textcolor{blue}{L7 - "a" here should be removed. L17 - "suits best" should be "best suits". L23 - "allowing to" should be "allowing users to". L28 - "suit" should be "suite".}}

I've fixed these typos.

\item{\textcolor{blue}{L54 - Technically a principal coordinates analysis *is* classic multidimensional scaling (CMDS). The author may also want to cite use of non-metric multidimensional scaling (e.g., Liow 2004, American Naturalist, 164, 431–).}}
\label{PCO_cites}

I have changed this sentence to:

\textit{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (Lloyd 2016 and references therein; e.g. in Close et al. 2015), or as outputs from an ordination, whether it being a principal components analysis (PCA, Hotelling 1933; e.g. in Zelditch et al. 2012), a metric scaling (PCO, PCoA, Torgerson 1958; e.g. in Brusatte et al. 2008) or a non-metric scaling (MDS, NMDS, Shepard 1962; e.g. in Liow 2004; Donohue et al. 2013).} lines @@@

@@

\item{\textcolor{blue}{L80 - I'm not sure I would use the term "enforce" here. I think many of us write packages that serve our own needs, but the great thing about languages like R is that it is flexible and code can be borrowed and appropriated etc. Essentially I think the author may unintentionally disparage other package authors here. Besides which, I do not feel that dispRity is competing with other packages, rather it is supplemental to them. I.e., to repeat my comment from above I see dispRity as sitting at the end of a range of data analysis pipelines with other packages sitting in front of it.}}

I've change this sentence to:

\textit{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}  lines @@@

@@

\item{\textcolor{blue}{L245-247 - I don't think these definitions for crown and stem are quite right. However, for the audience of MEE I suspect there is no need to define them so these parenthetic statements can just be cut.}}

I've removed the inaccurate definitions in parenthesis.

\item{\textcolor{blue}{L252 - It might be more helpful to define these using the phylogeny as this is a more generally applicable approach than manually delimiting row numbers (where human error can easily creep in or the inclusion of exclusion of data can mess things up). I.e., something that identifies descendants of the crown node and then a simple setdiff() between these and all possible names to get the non-crown (stem) taxa).}}

Great suggestion! I've added an utility function \texttt{crown.stem} allowing to separate stem and crown species.
The function is now briefly mentioned in the package description paper and detailed in the package itself.

@@

\begin{enumerate}

\item{\textcolor{blue}{The author presents a simple example and defers much of the many details users will encounter to his impermanent sources (vignettes, webpages), which I believe to be wise, as the R package will evolve.  I believe the purpose of a manuscript like this is illustrate its purpose and utility rather than serve as an instructional manual.  I believe the author has found the appropriate balance of instructional content between this manuscript and his other sources.}}

\item{\textcolor{blue}{The author acknowledged assistance with writing (suspect English is a second language) and the writing was great in the paper, but less so in the abstract.  Having someone double check the grammar in the abstract would be beneficial.  Also, line 116 refers to table ??.}}

\item{\textcolor{blue}{In the examples, replacing objects is not a good idea.  (This is also a problem in the help file examples in the R package.).  Remember that novice R users will use this software and might develop the habit of overwriting objects in the global environment, as updates.  It is possible to update objects with new names and retain the history during analyses, which is preferred in most cases and essential in others (for example, when a different analysis requires going back to a previously defined object.)  I recommend providing unique names.  For example, disparity is defined twice as an object on lines 255 and 261.  On line 261 it could be defined as DisparityB, and subsequently used without overwriting the object produced from a different function in line 255. Relatedly, one should not name objects the same name as functions.  This is simply bad news.  It is not something to encourage users to do.  Using "Disparity" rather than the function name, disparity, accomplishes the same thing without potentially causing problems down the line.  (The author will field requests for help from users, so this can be considered a pitfall avoided beforehand.)  This is discussed some here: https://stackoverflow.com/questions/6135868/in-r-what-exactly-is-the-problem-with-having-variables-with-the-same-name-as-ba}}

I have changed the variable names in the package description manuscript to not be redundant.
I've also changed the variable names in the package's help files to not override other functions/objects.

@@

\item \textcolor{blue}{2) Consider removing underdeveloped functions or arguments from your package until you have had sufficient time to test them.  Release them along with subsequent manuscripts that show you have tested type 1 error rates and statistical power.}

\item \textcolor{blue}{3) Or (and less recommended), put very stark and clear warnings that statistical results produced by some of these functions have not been tested and the user bares some risk making inferences from them.}

\end{itemize}

The ANOVA problem is totally a garbage-in-garbage-out problem and I agree with this reviewer that this might be a serious problem down the line.

Unfortunately, I do not have any proper solution to propose to this problem and I guess multidimensional distributions comparisons methods needs yet to be developed!

The reason for these crazy high degrees of freedom come from the bootstrapping pseudo-replication: effectively, the test data contains all the bootstrap data.
One solution is to use the concatenate = FALSE option in test.dispRity, performing the test for each bootstrap replicate rather than pooling the data.
This methods outputs a distribution of test that can be used to assess the type I error.
However, the pooling of the p-value is not satisfactory either.

Some methods exists such as NPMANOVA (from vegan::adonis) but is performed on the wrong kind of matrix.

One of the reasons for this functionality in the package is to make the test.dispRity function modular like the other main functions: i.e. it \textit{can} apply a \textit{lm} class function to the data but the reason for doing that might be dubious (or not, if the users don't bootstrap the matrix for example).

%TG: I therefore removed the example in the manuscript and the manual to not mislead users into running flawed analysis. However, I've left the architecture implemented for adventurous (or clever) users.
% Remove the ANOVA example
% Remove mentions to it on bootstrapped data in the manuals
% Replace by more savy example
% Allow to use NPMANOVA instead

@@

\textcolor{blue}{If you want to aim for a more general audience, consider rewriting the manuscript in more general terminology; samples in rows, variables in columns, for example.
Perhaps include a table to relate terminology used in morphometry with that more commonly encountered in community ecology.
Consider adding an ecological example and removing, if length is an issue, content on the manual and vignette --- we don't really need a paragraph on that, just a sentence to highlight it exists.}}

% %TG: adding a table of terms correspondance
% %TG: adding an ecology example (removing manual and vignette description)

\item{\textcolor{blue}{The referencing is poor; why cite Close et al for the concept of pairwise distances?
I'm pretty sure PCO existed before Brusatte et al, and PCA certainly before 2012, etc.
Throughout I was frustrated with this; the referencing ignores a century of work outside this specific field.
I appreciate there merit in citing relevant literature for your field, but that doesn't absolve one of a duty to accurately attribute ideas to those that came up with them.
If you are unfamiliar with developments in ideas beyond your particular field, I would suggest Legendre \& Legendre's *Numerical Ecology* textbook as a good source for original work relevant to what is presented here.}}

See my response to Reviewer 2 - Minor suggestion \ref{PCO_cites}.

### 2017-11-20 paper/dispRity_paper_highlighted.tex
Responses to reviewer 3

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix \citep[][and references therein]{lloyd2016estimating}, or as outputs from a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, or a multidimensional scaling \citep[MDS;][]{DonohueDim}, among others.
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.

@@

% Problems with this variety
In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric}, package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric and then enforce this on users by allowing little to no flexibility.
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
\hl{as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)}
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
\hl{This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).}
This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!

@@

\noindent These two groups can be defined as follows (where the integers are the row number of the species in the matrix):

\noindent \texttt{> mammal\_groups <- list(\textquotedbl crown\textquotedbl = c(16, 19:41, 45:50), \textquotedbl stem\textquotedbl = c(1:15, 17:18, 42:44))}

@@

\noindent \texttt{> subsets <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
\noindent \texttt{> bootstraps <- boot.matrix(subsets)}\\
\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

@@

@@

\noindent \texttt{> plot(disparity, type = \textquotedbl continuous\textquotedbl)}

@@

\noindent \texttt{> anova(test.dispRity(disparity, test = lm, comparisons = \textquotedbl all\textquotedbl))}

>>>>

% Variety of multidimensional spaces, metrics and tests
In all these analysis, each set of multivariate traits forms a multidimensional space.
\hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
\hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
\citealt{lloyd2016estimating}
\hl{and references therein; e.g. in}
\citealt{Close2015}
\hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
\citealt{PCA}
\hl{; e.g. in}
\citealt{zelditch2012geometric}
\hl{), a metric scaling (PCO, PCoA,}
\citealt{PCO}
\hl{; e.g. in}
\citealt{Brusatte12092008}
\hl{) or a non-metric scaling (MDS, NMDS,}
\citealt{MDS}
\hl{; e.g. in}
\citealt{Liow2004,DonohueDim}
\hl{).}
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.

@@

% Problems with this variety
In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods \citep[such as][for biogeography]{matzke2013biogeobears}.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,adams2013geomorph,Claddis,adams2017geometric},
\hl{package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best it their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.}
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
\hl{as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)}
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
\hl{This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).}
This can lead to inappropriate analyses by users confined by the existing software implementations. %Oh no!

@@

\noindent These two groups can be defined using one the package's utility functions, \texttt{crown.stem} that selects :

\noindent \texttt{> mammal\_groups <- crown.stem(BeckLee\_tree)}

@@

\noindent \texttt{> \hl{time\_subsets} <- time.subsets(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = \textquotedbl discrete\textquotedbl)}\\
\noindent \texttt{> \hl{bootstraps2} <- boot.matrix(\hl{time\_subsets})}\\
\noindent \texttt{> \hl{disparity2} <- dispRity(\hl{bootstraps2}, metric = c(sum, variances))}\\

@@

\noindent \texttt{> summary(\hl{disparity2})}

@@

\noindent \texttt{> plot(\hl{disparity2}, type = \textquotedbl continuous\textquotedbl)}

@@

\noindent \texttt{> anova(test.dispRity(disparity2, test = lm, comparisons = \textquotedbl all\textquotedbl))}
